---
title: "Apple Silicon power and frequency analysis"
format: 
 html:
    toc: true
    embed-resources: true
    df-print: paged
---

## Data overview

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(colorspace)

# load the data
data <- imap(list.files("results", "\\.csv$", full.names = TRUE), ~ {
  tibble(
    .source = ..2,
    read_csv(..1,  show_col_types = FALSE)
  )
}) %>% 
bind_rows() %>%
# single- vs. multi-core results (by number of threads)  
mutate(
  type = ifelse(n_distinct(thread_id) == 1, "single", "multi"),
  .by  = .source,
  .after = .source
) %>%
# P-core counters
mutate(
  # P-core frequency and power (account for zero denominator)
  p_freq = ifelse(p_time > 0.01, p_cycles / p_time / 1e9, 0),
  p_power = ifelse(p_time > 0.01, p_energy / p_time, 0),
  # relative thread time spend on P-core 
  p_usage = p_time/(p_time + e_time),
  .before = p_cycles
) %>%
# E-core counters
mutate(
  # E-core frequency and power  (account for zero denominator)
  e_freq = ifelse(e_time > 0.01, e_cycles / e_time / 1e9, 0),
  e_power = ifelse(e_time > 0.01, e_energy / e_time, 0),
  # relative thread time spend on E-core 
  e_usage = e_time/(p_time + e_time),
  .before = e_cycles
) %>%
# additional aggregated stats per thread
mutate(
  # total power
  power = (p_energy + e_energy)/(p_time + e_time),
  # throughtput (items/sec — both P- and E-cores) 
  items_tp = items / (p_time + e_time),
  # which cores did the thread run at
  cores = {
    P <- p_usage > 0.05
    E <- e_usage > 0.05
    case_when(P & E ~ "P,E", P ~ "P", E ~ "E")
  },
  .after = powermode
) %>%
# device detection
mutate(
  device = case_when(
    # A13
    device == "iPhone12,1"     ~ "A13",
    device == "iPhone12,8"     ~ "A13",
    
    # A14
    device == "iPhone13,3"     ~ "A14",
    
    # A15
    device == "iPad14,1"       ~ "A15", # iPad Mini 6
    device == "iPhone14,2"     ~ "A15",
    
    # A16
    device == "iPhone15,2"     ~ "A16",
    device == "iPhone15,3"     ~ "A16",
    
    # A17 Pro
    device == "iPhone16,1"     ~ "A17 Pro",
    
    # M1 family
    device == "MacBookPro18,2" ~ "M1", # Max
    device == "MacBookPro18,3" ~ "M1", # Pro
    device == "Mac13,2"        ~ "M1", # Ultra
    
    # M2 family
    device == "Mac14,5"       ~ "M2", # Max
    device == "Mac14,13"      ~ "M2", # Max
    device == "Mac14,6"      ~ "M2", # Max
    device == "Mac14,14"      ~ "M2", # Ultra
    TRUE                      ~ "unknown"
  )
) 

# guard agains unknown device
stopifnot(!any(data$device == "unknown"))
```

The test data is shown in the table below. We have the following columns

- **.source** the test run ID
- **type** singe- or multi-core run
- **sample** test iteration
- **thread_id** thread (unique across samples)
- **device** the device type
- **powermode** high vs. low power
- **power** total power usage of the thread per sample
- **items_tp** work throughput per thread per the sample (items/sec)
- **cores** types of cores the thread was running on
- **p_freq** average frequency of the P-core while running the thread (GHz)
- **p_power** average power draw of the P-core while running the thread (watt)
- **p_usage** relative amount of time the thread spend running on a P-core
- **p_cycles** number of cycles spent on a P-core
- **p_time** amount of time spend on a P-core
- **p_energy** energy used by P-core running the thread (J)
- **e_...** same as above, but for E-cores
- **items** number of work items processed by the thread for this sample

```{r}
data
```


## Power and frequency by device

Power draw distribution across devices in high power mode

```{r}
ggplot(
  data %>% 
    filter(powermode == "high") %>% 
    # remove multi-core for Macs, since it makes the gaph hard to read
    filter(startsWith(device, "A") | type == "single") %>%
    # sum the total power per sample (to get proper multi-core power usage)
    summarize(power = sum(power), .by = c(.source, sample, type, device))
) +
geom_violin(aes(x = device, y = power, fill = type), alpha=0.5, scale = "width") + 
ggtitle("Power consumption distribution by device (high power mode)") +
xlab("Average power draw, watts")
```

Power draw distribution across devices in low power mode

```{r}
ggplot(
  data %>% 
    filter(powermode == "low") %>% 
    # remove multi-core for Macs, since it makes the gaph hard to read
    filter(startsWith(device, "A") | type == "single") %>%
    # sum the total power per sample (to get proper multi-core power usage)
    summarize(power = sum(power), .by = c(.source, sample, type, device))
) +
geom_violin(aes(x = device, y = power, fill = type), alpha=0.5, scale = "width") + 
ggtitle("Power consumption distribution by device (low power mode)") +
xlab("Average power draw, watts")
```

P-thread power/frequency relationship by device (power curve). We only consider thread samples where at least 10% of time was spend on a P-core to avoid obviously unreliable numbers. The data for A-series shows considerable amount of variation since we combine together single- and multi-core mode, as well as high- and low-power mode (and the phones do throttle their frequency over time, giving us a larger frequency range within these scenarios). The M-series appear more grouped because they are less prone to frequency adjustments over time due to their higher thermal dissipation capability. 

```{r}
ggplot(filter(data, p_power > 0.1)) + 
  geom_point(aes(x = p_freq, y = p_power, color = device), size = 0.75, alpha = 0.5) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  ggtitle("P-core power curve by device") +
  ylab("Thread power draw, watts") + 
  xlab("CPU frequency, GHz")
```

Same, but for E-cores


```{r}
ggplot(filter(data, e_power > 0.1)) + 
  geom_point(aes(x = e_freq, y = e_power, color = device), size = 0.75, alpha = 0.5) +
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  ggtitle("E-core power curve by device") +
  ylab("Thread power draw, watts") + 
  xlab("CPU frequency, GHz")
```

## Work throughput and energy efficiency

Please take all of this with a grain of salt since the test is very simplistic and won't account for IPC differences between the micro-architectures. This is just about rough behavior. Work efficiency shoudl be measured on concrete workloads. 


Work throughput (items/sec) for each device (high power, single-core results). Note that there is no good way to know how much work was done
on P-cores and E-cores, so we just put this together. The different proportions of core involvement across threads produces the "cloud" effect. 

```{r}
ggplot(filter(data, powermode == "high", type == "single")) + 
  geom_point(
    aes(x = items/(p_time + e_time), y = p_power + e_power, color = device), 
    size = 0.75, 
    alpha = 0.75
  ) + 
  scale_color_discrete_qualitative("Set2") + 
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  ggtitle("Work throughtput by power usage (high power, single-core)") +
  ylab("Thread power (P+E combined), watts") + 
  xlab("Work throughtput, items/s") 
```

Energy (in J) used to perform the work

```{r}
ggplot(filter(data, powermode == "high", type == "single")) + 
  geom_point(
    aes(x = items, y = p_energy + e_energy, color = device), 
    size = 0.75, 
    alpha = 0.75
  ) + 
  scale_color_discrete_qualitative("Set2") + 
  scale_x_continuous(n.breaks = 10) +
  scale_y_continuous(n.breaks = 10) +
  ggtitle("Energy usage and work per thread (high power, single-core)") +
  xlab("Total work performed by thread (items) ") + 
  ylab("Energy used (J)") 
```

## Power curve forecasting

Attempt to estimate how the A17 power curve will continue beyond the observed range. We use the 
constrained regression algorithms from package `colf` to force the coefficients to be non-negative (as 
increasing frequency cannot lower power consumption). Please take this with a big grain of  salt, as this is likely overfitting the data. 


```{r}
# P-core data for A17 Pro
a17_data <- data %>%
  filter(device == "A17 Pro", p_usage > 0.2) %>% 
  mutate(p_freq2 = p_freq^2, p_freq3 = p_freq^3, p_freq4 = p_freq^4, p_freq5 = p_freq^5, p_freq6 = p_freq^6)

# fit the polynomial using constrained linear regression
m1 <- colf::colf_nlxb(p_power ~ 1 + p_freq, a17_data, lower = c(-Inf, 0))
m2 <- colf::colf_nlxb(p_power ~ 1 + p_freq + p_freq2, a17_data, lower = c(-Inf, 0, 0))
m3 <- colf::colf_nlxb(p_power ~ 1 + p_freq + p_freq2 + p_freq3, a17_data, lower = c(-Inf, 0, 0, 0))
m4 <- colf::colf_nlxb(p_power ~ 1 + p_freq + p_freq2 + p_freq3 + p_freq4, a17_data, lower = c(-Inf, 0, 0, 0, 0))
m5 <- colf::colf_nlxb(p_power ~ 1 + p_freq + p_freq2 + p_freq3 + p_freq4 + p_freq5, a17_data, lower = c(-Inf, 0, 0, 0, 0, 0))
m6 <- colf::colf_nlxb(p_power ~ 1 + p_freq + p_freq2 + p_freq3 + p_freq4 + p_freq5 + p_freq6, a17_data, lower = c(-Inf, 0, 0, 0, 0, 0, 0))


# there is not much difference beyond fourth-degree polynomial
a17_curve <- m4



# setup predicted data
a17_prediction <- tibble(
  p_freq = seq(1, 5, by = 0.05), 
  p_freq2 = p_freq^2, 
  p_freq3 = p_freq^3, 
  p_freq4 = p_freq^4,
  p_freq5 = p_freq^5,
  p_freq6 = p_freq^6
)
a17_prediction$p_power <- predict(m4, a17_prediction)

ggplot(a17_prediction) + 
  geom_path(aes(x = p_freq, y = p_power), color = "cyan", linetype = "dashed") +
  # add the actual data
  geom_point(aes(x = p_freq, y = p_power), size = 0.75, alpha = 0.75, data = a17_data) +
  ggtitle("Predicted frequency curve for A17 Pro") +
  ylab("Thread power draw, watts") + 
  xlab("CPU frequency, GHz")
```

Compare this to A15/M2

```{r}
# P-core data for A15 
a15_data <- data %>%
  filter(device == "A15", p_usage > 0.2) %>% 
  mutate(p_freq2 = p_freq^2, p_freq3 = p_freq^3, p_freq4 = p_freq^4,p_freq5 = p_freq^5, p_freq6 = p_freq^6)



# fit the polynomial using constrained linear regression
m1 <- colf::colf_nlxb(p_power ~ 1 + p_freq, a15_data, lower = c(-Inf, 0))
m2 <- colf::colf_nlxb(p_power ~ 1 + p_freq + p_freq2, a15_data, lower = c(-Inf, 0, 0))
m3 <- colf::colf_nlxb(p_power ~ 1 + p_freq + p_freq2 + p_freq3, a15_data, lower = c(-Inf, 0, 0, 0))
m4 <- colf::colf_nlxb(p_power ~ 1 + p_freq + p_freq2 + p_freq3 + p_freq4, a15_data, lower = c(-Inf, 0, 0, 0, 0))
m5 <- colf::colf_nlxb(p_power ~ 1 + p_freq + p_freq2 + p_freq3 + p_freq4 + p_freq5, a15_data, lower = c(-Inf, 0, 0, 0, 0, 0))
m6 <- colf::colf_nlxb(p_power ~ 1 + p_freq + p_freq2 + p_freq3 + p_freq4 + p_freq5 + p_freq6, a15_data, lower = c(-Inf, 0, 0, 0, 0, 0, 0))


# the sixth degree polynomial offers the best fit and predicts M2 data well!
a15_curve <- m6


# setup predicted data
a15_prediction <- tibble(
  p_freq = seq(1, 5, by = 0.05), 
  p_freq2 = p_freq^2, 
  p_freq3 = p_freq^3, 
  p_freq4 = p_freq^4,
  p_freq5 = p_freq^5,
  p_freq6 = p_freq^6
)
a15_prediction$p_power <- predict(a15_curve, a15_prediction)


ggplot(a15_prediction) + 
  geom_path(aes(x = p_freq, y = p_power), color = "cyan", linetype = "dashed") +
  # add the actual data
  geom_point(aes(x = p_freq, y = p_power), size = 0.75, alpha = 0.75, data = {
    filter(data, device %in%  c("A15", "M2"), p_usage > 0.2)
  }) +
  # add the A17 curve for comparison
  geom_path(aes(x = p_freq, y = p_power), color = "grey", linetype = "dashed", data = a17_prediction) +
  geom_label(aes(x = p_freq, y = p_power), label = "a17", data = filter(a17_prediction, p_freq == 4.5)) + 
  ggtitle("Predicted frequency curve for A17 Pro") +
  ylab("Thread power draw, watts") + 
  xlab("CPU frequency, GHz")
```

